Dynamics of Colombo's Top: Tidal Dissipation and Resonance Capture

### 2021-05-17 writeups/Su_weak_tides.tex
05/18 meeting

We next seek to characterize the stability of small perturbations about each of
the CSs in the presence of the weak alignment torque. We can linearize
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) about a shifted CS, yelding
\begin{align}
    \rd{}{t}\begin{bmatrix}
        \theta\\ \phi
    \end{bmatrix} &= \begin{bmatrix}
        -\frac{\cos \theta}{t_{\rm al}} &
        -g\sin I \cos \phi \\
        \alpha \sin \theta + g\frac{\sin I \cos \phi}{\sin^2\theta} &
        0
    \end{bmatrix}_{\rm cs}\begin{bmatrix}
        \Delta \theta \\ \Delta \phi
    \end{bmatrix},\label{eq:dsdt_hessian}
\end{align}
where the cs subscript indicates evaluating at a CS, $\Delta \theta = \theta -
\theta_{\rm cs}$, and $\Delta \phi = \phi - \phi_{\rm cs}$. The eigenvalues
$\lambda$ of Eq.~\eqref{eq:dsdt_hessian} satisfy the equation
\begin{equation}
    0 = \p{\lambda + \frac{\cos \theta_{\rm cs}}{t_{\rm al}}}\lambda
        - \lambda_0^2,\label{eq:lambda_orig}
\end{equation}
where
\begin{equation}
    \lambda_0^2 \equiv \p{\alpha
        \sin \theta_{\rm cs} + g\sin I \csc^2\theta_{\rm cs}\cos \phi_{\rm cs}}
            \p{- g \sin I \cos \phi_{\rm cs}}\label{eq:def_l0_sq}.
\end{equation}
When $t_{\rm al}$ is large, we can simplify Eq.~\eqref{eq:lambda_orig} to
\begin{equation}
    \lambda \approx -\frac{\cos \theta_{\rm cs}}{t_{\rm al}}
        \pm \sqrt{\lambda_0^2}. \label{eq:lambda_approx}
\end{equation}
The stability depends only on the real part of $\lambda$ in
Eq.~\eqref{eq:lambda_approx}. $\lambda_0^2$ is a generalization of Eq.~(A4)
in Paper I and generally has the same behavior: it is negative for CSs
1--3 and positive for CS4, as shown in Fig.~\ref{fig:lambda_full}. Thus, CS4 is
always unstable, as there will always be at least one positive solution for
$\lambda$, and the stability of CSs 1--3 are solely determined by the sign of
$\cos \theta_{\rm cs}$. Using Fig.~\ref{fig:cs_locs}, we conclude that CS1 and
CS2 are stable and attracting while trajectories near CS3 are driven away by the
alignment torque. These calculations justify results long used in CS literature
\citep[e.g.][]{ward1975tidal, fabrycky_otides}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_lambdas_full.png}
    \caption{$\lambda_0^2$ (Eq.~\ref{eq:def_l0_sq}) as a function of $\eta$ for
    the four CSs, for three different values of the shift in $\phi_{\rm cs}$
    (e.g.\ for $\Delta \phi_{\rm cs} = 60^\circ$, $\phi_{\rm cs} = 120^\circ$
    for CS2 and $\phi_{\rm cs} = 60^\circ$ for CSs 1, 3, and 4). The values of
    $\Delta \phi_{\rm cs}$ are labeled. }\label{fig:lambda_full}
\end{figure}

@@

Using this result, we can then identify the long-term equilibria of the system
when $\Omega_{\rm s}$ is also evolving: these equilibria of the system must both
satisfy $\dot{\Omega}_{\rm s} = 0$ and be a CS that is stable in the presence of
tidal dissipation (i.e.\ satisfying $\rdil{\uv{s}}{t} = 0$).
The panels of Fig.~\ref{fig:6equils} describes the evolution of
Eqs.~(\ref{eq:dsdt_tide}--\ref{eq:dWsdt_tide}) qualitatively in the coordinates
$(\Omega_{\rm s}, \theta)$, along with the locations of CS1 and CS2, for a few
different system parameters. The circled points in the panels of
Fig.~\ref{fig:6equils} satisfy the criteria to be long-term equilibria, and we
call them \emph{tidal Cassini Equilibria} (tCE). We number tCE1 and tCE2 the tCE
that are in CSs 1 and 2 respectively. The obliquities of the tCE depend on the
system architecture, which can be quantified using the parameter
\begin{align}
    \eta_{\rm sync} &\equiv \big(\eta\big)_{\Omega_{\rm s} = n}
        = \eta \frac{\Omega_{\rm s}}{n}\nonumber\\
        &= \frac{1}{2}\frac{k}{k_{\rm q}}
            \frac{m_{\rm p}m}{M_\star^2}
            \p{\frac{a}{a_{\rm p}}}^3 \p{\frac{a}{R}}^3\cos(I).
            \label{eq:def_etasync}
\end{align}
The tCE obliquity as a function of $\eta_{\rm sync}$ are shown for $I =
20^\circ$ in the top panel of Fig.~\ref{fig:probs20} and for $I = 5^\circ$ in
the top panel of Fig.~\ref{fig:probs5}. In fact, an analytical expression for
the tCE2 obliquity for $\eta_{\rm sync} \ll \eta_{\rm c}$ can be computed using
Eqs.~(\ref{eq:weaktide_dWszero}--\ref{eq:def_etasync}) and that $\cos \theta_2
\approx \eta \cos I$ to yield:
\begin{align}
    \cos \theta_{\rm tCE2} &\approx \sqrt{\frac{\eta_{\rm sync} \cos I}{2}},
        &
    \frac{\Omega_{\rm s, tCE2}}{n} &\approx
        \sqrt{2\cos I \eta_{\rm sync}}. \label{eq:def_tce2_approx}
\end{align}
This approximation for $\theta_{\rm tCE2}$ is shown as the black dashed line in
the top panels of Fig.~\ref{fig:probs20} and~\ref{fig:probs5}, where good
agreement is observed.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/6equils.png}
    \caption{Schematic depiction of the effect of tidal friction on the planet's
    spin for $I = 20^\circ$ and three values of $\eta_{\rm sync}$ (labeled). The
    black and blue lines denote where the tidal $\dot{\Omega}_{\rm s}$ and
    $\dot{\theta}$ change signs
    (Eqs.~\ref{eq:weaktide_dqzero}--\ref{eq:weaktide_dWszero}). The orange and
    green lines give the CS1 and CS2 obliquities respectively, which are the two
    CSs that are stable under the effect of tidal dissipation. Note that when
    $\dot{\theta}_{\rm tide} > 0$, CS1 becomes unstable, denoted by the dashed
    orange line. The points that are both CSs and satisfy $\dot{\Omega}_{\rm s}
    = 0$ are the tidal Cassini Equilibria (tCE), which are circled and labeled.
    }\label{fig:6equils}
\end{figure}

There are two important conditions that can change the existence and stability
of the tCE\@. First, if $\eta_{\rm sync} > \eta_{\rm c}$, then tCE1 will not
exist, as can be seen in the bottom panel of
Fig.~\ref{fig:6equils}\footnote{Strictly speaking, $\eta_{\rm sync}$ can be
slightly smaller than $\eta_{\rm c}$, as the planet's spin is slightly
subsynchronous at tCE1 when $\eta \approx \eta_{\rm c}$, see the middle panel of
Fig.~\ref{fig:6equils}.}. Secondly, tCE2 may not be stable if the tidal phase
shift is too large. Applying the results of Section~\ref{ss:tidal_equils}, we
find that tCE2 is stable as long as $t_{\rm s} \geq t_{\rm s, c}$ where
\begin{equation}
    t_{\rm s, c} \equiv \frac{\sin \theta_{\rm tCE2}}{\abs{g} \sin
            I}\p{\frac{2n}{\Omega_{\rm s, tCE2}} - \cos \theta_{\rm tCE2}}.
\end{equation}
The tCE2 subscripts denote evaluation at tCE2. When $\eta_{\rm sync} \ll
\eta_{\rm c}$, we can use Eq.~\eqref{eq:def_tce2_approx} to further simplify:
\begin{equation}
    t_{\rm s, c}\approx \frac{1}{\abs{g}\sin I}\sqrt{\frac{2}{\eta_{\rm sync}
        \cos I}}. \label{eq:def_ts_crit}
\end{equation}

@@

To validate the accuracy of Eq.~\eqref{eq:def_pc_weaktide}, we can compare with
direct numerical integration of
Eqs.~(\ref{eq:dsdt_rot},~\ref{eq:dsdt_tide}--\ref{eq:dWsdt_tide}) for many
initial conditions while evaluating $P_{\rm III \to II}$ (and thus also
obtaining $P_{\rm III \to I} = 1 - P_{\rm III \to II}$) for each simulation at
the moment it encounters the separatrix, if it does so. If the theory is
correct, the total numbers of systems converging to each of tCE1 and tCE2 should
be equal to those predicted by the sums of the calculated probabilities. In
Fig.~\ref{fig:pc_fits_0_06}, we show the agreement of this semi-analytic
procedure with the numerical results displayed earlier in the right panel of
Fig.~\ref{fig:Hhists_0_06}. Good agreement is observed.
Figs.~\ref{fig:pc_fits_0_20} shows the same for Figs.~\ref{fig:Hhists_0_20}.
Mostly satisfactory agreement is observed. Thus, we conclude that the outcomes
of separatrix encounter are accurately predicted by
Eq.~\eqref{eq:def_pc_weaktide}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5pc_fits0_06_20.png}
    \caption{Comparison of the fraction of systems converging to tCE2 obtained
    via numerical simulation (red dots) and obtained via a semi-analytic
    calculation (blue line) for $\eta_{\rm sync} = 0.06$ and $I = 20^\circ$ (see
    right panel of Fig.~\ref{fig:Hhists_0_06}). The semi-analytic calculation is
    performed by numerically integrating
    Eqs.~(\ref{eq:dsdt_rot},~\ref{eq:dsdt_tide}--\ref{eq:dWsdt_tide}) on a grid
    of initial conditions uniform in $\cos \theta_{\rm i}$ and $\phi_{\rm i}$
    until the system reaches the separatrix, then calculating the probability of
    reaching tCE2 for each integration using
    Eq.~\eqref{eq:def_pc_weaktide}.}\label{fig:pc_fits_0_06}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5pc_fits0_20_20.png}
    \caption{Same as Fig.~\ref{fig:pc_fits_0_06} but for $\eta_{\rm sync} =
    0.2$, corresponding to the right panel of
    Fig.~\ref{fig:Hhists_0_20}.}\label{fig:pc_fits_0_20}
\end{figure}

>>>>

We next seek to characterize the stability of small perturbations about each of
the CSs in the presence of the weak alignment torque. We can linearize
Eqs.~(\ref{eq:dqdt_toy}--\ref{eq:dfdt_toy}) about a shifted CS, yelding
\begin{align}
    \rd{}{t}\begin{bmatrix}
        \Delta \theta\\ \Delta \phi
    \end{bmatrix} &= \begin{bmatrix}
        -\frac{\cos \theta}{t_{\rm al}} &
        -g\sin I \cos \phi \\
        \alpha \sin \theta + g\frac{\sin I \cos \phi}{\sin^2\theta} &
        0
    \end{bmatrix}_{\rm cs}\begin{bmatrix}
        \Delta \theta \\ \Delta \phi
    \end{bmatrix},\label{eq:dsdt_hessian}
\end{align}
where the cs subscript indicates evaluating at a CS, $\Delta \theta = \theta -
\theta_{\rm cs}$, and $\Delta \phi = \phi - \phi_{\rm cs}$. The eigenvalues
$\lambda$ of Eq.~\eqref{eq:dsdt_hessian} satisfy the equation
\begin{equation}
    0 = \p{\lambda + \frac{\cos \theta_{\rm cs}}{t_{\rm al}}}\lambda
        - \lambda_0^2,\label{eq:lambda_orig}
\end{equation}
where
\begin{equation}
    \lambda_0^2 \equiv \p{\alpha
        \sin \theta_{\rm cs} + g\sin I \csc^2\theta_{\rm cs}\cos \phi_{\rm cs}}
            \p{- g \sin I \cos \phi_{\rm cs}}\label{eq:def_l0_sq}.
\end{equation}
When $t_{\rm al}$ is large, we can simplify Eq.~\eqref{eq:lambda_orig} to
\begin{equation}
    \lambda \approx -\frac{\cos \theta_{\rm cs}}{t_{\rm al}}
        \pm \sqrt{\lambda_0^2}. \label{eq:lambda_approx}
\end{equation}
The stability depends only on the real part of $\lambda$ in
Eq.~\eqref{eq:lambda_approx}. $\lambda_0^2$ is a generalization of Eq.~(A4)
in Paper I and generally has the same behavior: it is negative for CSs
1--3 and positive for CS4, as shown in Fig.~\ref{fig:lambda_full}. Thus, CS4 is
always unstable, as there will always be at least one positive solution for
$\lambda$, and the stability of CSs 1--3 are solely determined by the sign of
$\cos \theta_{\rm cs}$. Using Fig.~\ref{fig:cs_locs}, we conclude that CS1 and
CS2 are stable and attracting while trajectories near CS3 are driven away by the
alignment torque. These calculations justify results long used in CS literature
\citep[e.g.][]{ward1975tidal, fabrycky_otides}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/99_misc/2_lambdas_full.png}
    \caption{$\lambda_0^2$ (Eq.~\ref{eq:def_l0_sq}) as a function of $\eta$ for
    the four CSs, for three different values of the shift in $\phi_{\rm cs}$
    (e.g.\ for $\Delta \phi_{\rm cs} = 60^\circ$, $\phi_{\rm cs} = 120^\circ$
    for CS2 and $\phi_{\rm cs} = 60^\circ$ for CSs 1, 3, and 4). The values of
    $\Delta \phi_{\rm cs}$ are labeled. }\label{fig:lambda_full}
\end{figure}

@@

Using this result, we can then identify the long-term equilibria of the system
when $\Omega_{\rm s}$ is also evolving: these equilibria of the system must both
satisfy $\dot{\Omega}_{\rm s} = 0$ and be a CS that is stable in the presence of
tidal dissipation (i.e.\ satisfying $\rdil{\uv{s}}{t} = 0$).
The panels of Fig.~\ref{fig:6equils} describes the evolution of
Eqs.~(\ref{eq:dsdt_tide}--\ref{eq:dWsdt_tide}) qualitatively in the coordinates
$(\Omega_{\rm s}, \theta)$, along with the locations of CS1 and CS2, for a few
different system parameters. The circled points in the panels of
Fig.~\ref{fig:6equils} satisfy the criteria to be long-term equilibria, and we
call them \emph{tidal Cassini Equilibria} (tCE). We number tCE1 and tCE2 the tCE
that are in CSs 1 and 2 respectively. The obliquities of the tCE depend on the
system architecture, which can be quantified using the parameter
\begin{align}
    \eta_{\rm sync} &\equiv \big(\eta\big)_{\Omega_{\rm s} = n}
        = \eta \frac{\Omega_{\rm s}}{n}\nonumber\\
        &= \frac{1}{2}\frac{k}{k_{\rm q}}
            \frac{m_{\rm p}m}{M_\star^2}
            \p{\frac{a}{a_{\rm p}}}^3 \p{\frac{a}{R}}^3\cos(I).
            \label{eq:def_etasync}
\end{align}
The tCE obliquity as a function of $\eta_{\rm sync}$ are shown for $I =
20^\circ$ in the top panel of Fig.~\ref{fig:probs20} and for $I = 5^\circ$ in
the top panel of Fig.~\ref{fig:probs5}. In fact, an analytical expression for
the tCE2 obliquity for $\eta_{\rm sync} \ll \eta_{\rm c}$ can be computed using
Eqs.~(\ref{eq:weaktide_dWszero}--\ref{eq:def_etasync}) and that $\cos \theta_2
\approx \eta \cos I$ to yield:
\begin{align}
    \cos \theta_{\rm tCE2} &\approx \sqrt{\frac{\eta_{\rm sync} \cos I}{2}},
        &
    \frac{\Omega_{\rm s, tCE2}}{n} &\approx
        \sqrt{2\cos I \eta_{\rm sync}}. \label{eq:def_tce2_approx}
\end{align}
This approximation for $\theta_{\rm tCE2}$ is shown as the black dashed line in
the top panels of Fig.~\ref{fig:probs20} and~\ref{fig:probs5}, where good
agreement is observed.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/6equils.png}
    \caption{Schematic depiction of the effect of tidal friction on the planet's
    spin for $I = 20^\circ$ and three values of $\eta_{\rm sync}$ (labeled). The
    black and blue lines denote where the tidal $\dot{\Omega}_{\rm s}$ and
    $\dot{\theta}$ change signs
    (Eqs.~\ref{eq:weaktide_dqzero}--\ref{eq:weaktide_dWszero}). The orange and
    green lines give the CS1 and CS2 obliquities respectively, which are the two
    CSs that are stable under the effect of tidal dissipation. Note that when
    $\dot{\theta}_{\rm tide} > 0$, CS1 becomes unstable, denoted by the dashed
    orange line. The points that are both CSs and satisfy $\dot{\Omega}_{\rm s}
    = 0$ are the tidal Cassini Equilibria (tCE), which are circled and labeled.
    \textcolor{red}{The various colored lines indicate sample evolutionary
    trajectories that are started at the same $\theta$ and $\Omega_{\rm s}$
    across all three panels, with different outcomes upon separatrix encounter.}
    }\label{fig:6equils}
\end{figure}

There are two important conditions that can change the existence and stability
of the tCE\@. First, if $\eta_{\rm sync} > \eta_{\rm c}$, then tCE1 will not
exist, as can be seen in the bottom panel of
Fig.~\ref{fig:6equils}\footnote{Strictly speaking, $\eta_{\rm sync}$ can be
slightly smaller than $\eta_{\rm c}$, as the planet's spin is slightly
subsynchronous at tCE1 when $\eta \approx \eta_{\rm c}$, see the middle panel of
Fig.~\ref{fig:6equils}.}. Secondly, tCE2 may not be stable if the tidal phase
shift is too large. Applying the results of Section~\ref{ss:tidal_equils}, we
find that tCE2 is stable as long as $t_{\rm s} \geq t_{\rm s, c}$ where
\begin{equation}
    t_{\rm s, c} \equiv \frac{\sin \theta_{\rm tCE2}}{\abs{g} \sin
            I}\p{\frac{2n}{\Omega_{\rm s, tCE2}} - \cos \theta_{\rm tCE2}}.
\end{equation}
The tCE2 subscripts denote evaluation at tCE2. When $\eta_{\rm sync} \ll
\eta_{\rm c}$, we can use Eq.~\eqref{eq:def_tce2_approx} to further simplify:
\begin{equation}
    t_{\rm s, c}\approx \frac{1}{\abs{g}\sin I}\sqrt{\frac{2}{\eta_{\rm sync}
        \cos I}}. \label{eq:def_ts_crit}
\end{equation}
\textcolor{red}{Note that this result differs in form from that given in
\citet{fabrycky_otides}. Our formulation is more appropriate and transparent
when the orbital precession of the inner planet is dominated by the outer planet
rather than the star.}

@@

To validate the accuracy of Eq.~\eqref{eq:def_pc_weaktide}, we can compare with
direct numerical integration of
Eqs.~(\ref{eq:dsdt_rot},~\ref{eq:dsdt_tide}--\ref{eq:dWsdt_tide}) for many
initial conditions while evaluating $P_{\rm III \to II}$ (and thus also
obtaining $P_{\rm III \to I} = 1 - P_{\rm III \to II}$) for each simulation at
the moment it encounters the separatrix, if it does so. If the theory is
correct, the total numbers of systems converging to each of tCE1 and tCE2 should
be equal to those predicted by the sums of the calculated probabilities. In
Fig.~\ref{fig:pc_fits_0_06}, we show the agreement of this semi-analytic
procedure with the numerical results displayed earlier in the right panel of
Fig.~\ref{fig:Hhists_0_06}. Good agreement is observed.
Figs.~\ref{fig:pc_fits_0_20} shows the same for Figs.~\ref{fig:Hhists_0_20}.
Mostly satisfactory agreement is observed. Thus, we conclude that the outcomes
of separatrix encounter are accurately predicted by
Eq.~\eqref{eq:def_pc_weaktide}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5pc_fits0_06_20.png}
    \caption{Comparison of the fraction of systems converging to tCE2 obtained
    via numerical simulation (red dots) and obtained via a semi-analytic
    calculation (blue line) for $\eta_{\rm sync} = 0.06$ and $I = 20^\circ$ (see
    right panel of Fig.~\ref{fig:Hhists_0_06}). The semi-analytic calculation is
    performed by numerically integrating
    Eqs.~(\ref{eq:dsdt_rot},~\ref{eq:dsdt_tide}--\ref{eq:dWsdt_tide}) on a grid
    of initial conditions uniform in $\cos \theta_{\rm i}$ and $\phi_{\rm i}$
    until the system reaches the separatrix, then calculating the probability of
    reaching tCE2 for each integration using
    Eq.~\eqref{eq:def_pc_weaktide}.
    \textcolor{red}{The green dashed line in the top panel shows the result of
    using the analytical expression (Eq.~\ref{eq:app_deltaK}) for $\Delta
    K_{\pm}$, and the bottom panel shows the distribution of values of
    $\eta_{\rm cross}$, the value of $\eta$ when a trajectory starting at
    $\theta_{\rm i}$ encounters the separatrix. }}\label{fig:pc_fits_0_06}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{../initial/1_weaktide/5pc_fits0_20_20.png}
    \caption{Same as Fig.~\ref{fig:pc_fits_0_06} but for $\eta_{\rm sync} =
    0.2$, corresponding to the right panel of
    Fig.~\ref{fig:Hhists_0_20}.}\label{fig:pc_fits_0_20}
\end{figure}
